---
title: "MATH 9102 - Probability and Statistical Inference Assignment - 2"
author: "Antonio Silva (D23129331@mytudublin.ie)"
output:
  html_document:
    toc: true
    toc_depth: 2
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
  pdf_document:
    toc: true
    toc_depth: '2'
---

```{r, echo = FALSE, results = "hide"}
if (!require('knitr',quietly = TRUE)) {
  install.packages('knitr', repos = 'http://cran.us.r-project.org')
}

if (!require('ggplot2', quietly = TRUE)) {
  install.packages('knitr', repos = 'http://cran.us.r-project.org')
}

if (!require('ggpubr', quietly = TRUE)) {
  install.packages('ggpubr', repos = 'http://cran.us.r-project.org')
}

if (!require('psych', quietly = TRUE, warn.conflicts = FALSE)) {
  install.packages('psych', repos = 'http://cran.us.r-project.org')
}

library(knitr)
library(ggplot2)
library(ggpubr)
library(psych)
```

# Question 1 [1 mark]
A study was done in which the high daily temperature and the number of traffic
accidents within the city were recorded. These sample data are shown as follows.
```
  temp <- c(91,56,75,68,50,39,98)
  accidents <- c(2,9,7,6,6,10,1)
  accTemp <- data.frame(temp, accidents)
```
Identify whether the data is positively correlated or negatively 
correlated using a scatter plot. 

## Answer 1

```{r, echo = TRUE, dev='png'}
temp <- c(91,56,75,68,50,39,98)
accidents <- c(2,9,7,6,6,10,1)
accTemp <- data.frame(temp, accidents)
ggplot(data = accTemp, aes(x = temp, y = accidents))  + 
  geom_point() + 
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE)
```

After we create the scatter plot and and the trend line we can tell that
the correlation is negative.

We can confirm calculating the correlation:
```{r, echo = TRUE, results = 'markdown'}
cor(accTemp$temp, accTemp$accidents)
```

# Question 2 [1 mark]

Following data gives the readings of sugar level of 5 diabetic patient 
before and after taking insulin. Using 1% level of significance test whether
insulin has reduced sugar level.
```
  before <- c(350,400,250,200,180)
  after <- c(200,300,200,150,120)
```

## Answer 2

Because we are talking about related samples where the individual participants
in single group measured at two different points the most suitable test is
the *Paired samples t-test*.

For the test we have the following Hypothesis: 
* Null hypothesis: The insulin did not changed the sugar levels;
* Alternative hypothesis: The insulin dropped the sugar levels.

So we will use the mean, because the measurements are taken from the same 
population, then they should fairly similar means.

The significance is 1%. 

$$
\begin{cases}
        H_0: \mu_d \ge 0 \ or \ \mu_{after} - \mu_{before} \ge 0 \\
        H_a: \mu_d < 0 \ or \ \mu_{after} - \mu_{before} < 0  
    \end{cases} 
$$

$$\alpha = 0.01$$

In R we can execute the test using the `t.test` function.
This is a one-tail test from the left side.

So the Confidence Interval is $CI = 1 - \alpha = 0.99$

```{r, echo = TRUE}
  before <- c(350,400,250,200,180)
  after <- c(200,300,200,150,120)
  t.test(after, before, paired = TRUE, alternative = "less", conf.level = 0.99)
```

So we got the t-value of $-4.2401$ which is not in the $99\%$ region of acceptance: 
$[-3.747, \infty[$. So it is safe to say that we *Reject the null hypothesis*
and conclude that the *insuline dropped the sugar levels*.

# Question 3 [2 marks]
The following dataset includes the score of students in mathematics and 
statistics. Justify with R code suitable correlation test. [Hint: Check for
normality and test the data for correlation]

```
  maths <- c(50,54,56,59,60,62,61,65,67,71,71,74)
  stat <- c(22,25,34,28,26,30,32,30,28,34,36,40)
  scoreSet <- data.frame(maths, stat)
```

## Answer 3 [2 marks]

First we start inspecting the both dataset to see if they belong to a normal
distribution. We can start by defining the data and create a Q-Q plot. 
As quoted in the labs of week 6:

> A Q-Q plot is a scatterplot created by plotting two sets of quantiles agains
> one another. If the both sets of quantiles come from the same distribution,
> we will see the points forming a line.

So to verify we can draw a Q-Q plot against the normal distribution first.

First we introduce the data
```{r, echo = TRUE}
maths <- c(50,54,56,59,60,62,61,65,67,71,71,74)
stat <- c(22,25,34,28,26,30,32,30,28,34,36,40)
scoreSet <- data.frame(maths, stat)
```

And we create a Q-Q plot for each dataset (maths and stat)

```{r, echo = TRUE, dev='png'}
ggqqplot(maths, ylab = "Math Quantiles", xlab="Normal Quantiles")
```

```{r, echo = TRUE, dev='png'}
ggqqplot(stat, ylab = "Stat Quantiles")
```

By the charts we can say that both datasets are close to a normal distribution,
but we can't assume that follows it.

To make sure its better to conduct a statistical normality test.
So we will run *Shapiro-Wilk test* against the datasets.

For the Shapiro Wilk tests we apply the following formula:
$$
W = \frac{(\sum a_ixi)^2}{\sum(x_i - \bar{x})^2}
$$
If the test is non-significant $(p > 0.05)$ it tells us 
that the distribution of the sample is not significantly 
different from a normal distribution. If,
however, the test is significant $(p < 0.05)$ then the 
distribution in question
is significantly different from a normal distribution (non-normal). 

```{r, echo = TRUE}
shapiro.test(maths)
shapiro.test(stat)
```

Because both p-values are significantly bigger than 0.05 (0.948, 0.9986)
we can consider that both datasets follow a normal distribution.

For that we will use the method Pearson to calculate the correlation
coefficient.

```{r, echo = TRUE}
cor.test(maths, stat, method='pearson')
```
So because we got a coefficient of $0.7835978$ we can tell that exists
a high correlation between maths ant statistics score.

You see the correlation by observing the scatter plot between the two.

```{r, echo = TRUE, dev='png'}
ggplot(data = scoreSet, aes(x = maths, y = stat))  + 
  geom_point() + 
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE)
```

# Question 4 [5 marks]
[Note: Save the read dataset in a variable called heartdisease. Do not change
the variable name. For ex: heardisease <- read.csv("heartdisease.data")]

Consider given dataset heartdisease.csv. Write code to do the 
following.

a) Assess the following variables for normality. [2 marks]
* Cholesterol (Chol)
* Blood Pressure (RestBP)
* MaxHR
b) Choose the test you think is correct based on your these variables
- choose either Pearson, Spearman / Kendall to investigate the
following: [3 marks]
* Relationship between cholesterol and blood pressure;
* Relationship between cholesterol and old peak.

## Answer 4

First we load the dataset
```{r, echo = TRUE}
heartdisease <- read.csv("heartdisease.csv")
```

Next we will test the three variables for normality with the Shapiro-Wilk test.

```{r, echo = TRUE}
shapiro.test(heartdisease$Chol)
shapiro.test(heartdisease$RestBP)
shapiro.test(heartdisease$MaxHR)
```

All the three values are bellow 0.05 so the data significantly deviate from a
normal distribution. For skew data we use Spearman/Kendal correlation.

For the relationship between cholesterol and blood pressure

```{r, echo = TRUE}
cor.test(heartdisease$Chol, heartdisease$RestBP, method = "spearman", exact = FALSE)
```
So the Correlation Coefficient value of 0.1358367 represents
a *weak positive monotonic correlation*.

The probability value is 0.018 and is less than 0.05 $\alpha$ threshold
meaning that the observation is statistically significant. 

So we assume that theres is a 
*weak correlation between cholesterol and blood pressure*. 
You can see the verify in the next plot.

```{r, echo = TRUE, dev='png'}
ggplot(data = heartdisease, aes(x = Chol, y = RestBP))  + 
  geom_point() + 
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE)
```

Relationship 
▪ Relationship between cholesterol and old peak

First let't test the `Oldpeak` variable for normality.

```{r, echo = TRUE}
shapiro.test(heartdisease$Oldpeak)
```
Again we get a really low p-value suggesting that we are not talking about data
not normally distributed. Again we will use the spearman test.

```{r, echo = TRUE}
cor.test(heartdisease$Chol, heartdisease$Oldpeak, method = "spearman", exact = FALSE)
```
The coefficient rho is 0.03435936 that represents a positive weak correlation
as presented in the chart.

```{r, echo = TRUE, dev='png'}
ggplot(data = heartdisease, aes(x = Chol, y = Oldpeak))  + 
  geom_point() + 
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE)
```

But because the p-value is above the assumed significant level $\alpha = 0.05$
we *fail to reject the null hypothesis*. 
We conclude that is no correlation/relationship between the cholesterol 
and old peak for a significance level of $0.05$.

# Question 5 [2 marks]
Investigate whether there are differences between people who survive
hepatitis and those who do not in each of the following variables.

* BILIRUBIN;
* SGOT.

[Note: Save the read dataset in a variable called hepatitis. Do not change
the variable name.
For ex: `hepatitis <- read.csv("hepatitis.data")`.
Dataset is available https://archive.ics.uci.edu/ml/datasets/Hepatitis.
Use the code below to add header.]
```
#To add headers
colnames(hepatitis) <- c("Class", "AGE", "SEX",
"STEROID", "ANTIVIRALS", "FATIGUE", "MALAISE",
"ANOREXIA", "LIVER BIG", "LIVER FIRM", "SPLEN
PALPABLE", "SPIDERS", "ASCITES",
"VARICES","BILIRUBIN","ALK PHOSPHATE",
"SGOT","ALBMIN", "PROTIME", "HISTOLOGY")
```

## Answer 5
First we need to load the data and add the headers. 

```{r, echo = TRUE}
hepatitis <- read.csv("hepatitis.data", quote="", na.strings = "?")

colnames(hepatitis) <- c(
  "Class", 
  "AGE", 
  "SEX",
  "STEROID", 
  "ANTIVIRALS", 
  "FATIGUE", 
  "MALAISE",
  "ANOREXIA", 
  "LIVER BIG", 
  "LIVER FIRM", 
  "SPLEN PALPABLE", 
  "SPIDERS", 
  "ASCITES",
  "VARICES",
  "BILIRUBIN",
  "ALK PHOSPHATE",
  "SGOT",
  "ALBMIN", 
  "PROTIME", 
  "HISTOLOGY")
```

First thing that we will do is look at the data and the structure

```{r, echo = TRUE}
str(subset(hepatitis, select = c("Class", "BILIRUBIN", "SGOT")))
```
So `class` is refered as int but in fact its a categorical class. 
We need to transform it in a character class type.

```{r, echo = TRUE}
hepatitis<- transform(hepatitis, Class = as.character(Class))
str(subset(hepatitis, select = c("Class", "BILIRUBIN", "SGOT")))
```
Now lets take a look at the data: 
```{r, echo = TRUE}
head(subset(hepatitis, select = c("Class", "BILIRUBIN", "SGOT")))
```
`Class` is a categorical variable that determines if the patient died or not:
 
 * 1 = DIE
 * 2 = LIVE
 It also does not has Missing values.
 
 `BILIRUBIN` is a continuous variable that represents a yellowish pigment that
 is made during the breakdown of red blood cells. It has 6 missing values.
 
 `SGOT` is a integer variable that represents SGOT is a protein made by the
 liver cells. It also has 4 missing values.
 
```{r, echo = TRUE}
sum(is.na(hepatitis$BILIRUBIN))
sum(is.na(hepatitis$SGOT))
```
 
Now that we analyze the data we can describe it: 

```{r, echo = TRUE}
table(hepatitis$Class)
psych::describe(hepatitis$BILIRUBIN)
psych::describe(hepatitis$SGOT)
``` 
So we have a sample where 32 individuals died and 122 survived.

Both `BILLRUBIN` and `SGOT` variables are positively skewed with 
a high kurtosis so it it suggest a non-normal distribution.

Now we will split the variables in survived not survived:

```{r, echo = TRUE}
hepatitis.BILLRUBIN.die = hepatitis[hepatitis$Class == 1, c("BILIRUBIN")]
hepatitis.BILLRUBIN.survived = hepatitis[hepatitis$Class == 2, c("BILIRUBIN")]

hepatitis.SGOT.die = hepatitis[hepatitis$Class == 1, c("SGOT")]
hepatitis.SGOT.survived = hepatitis[hepatitis$Class == 2, c("SGOT")]
``` 
To test the normality we will use the 
*Shapiro-Willis test for variables that die (32 size samples)*, and 
*Kolmogorov-Smirnov (122 sample size)*

```{r, echo = TRUE}
shapiro.test(hepatitis.BILLRUBIN.survived)
shapiro.test(hepatitis.SGOT.survived)

ks.test(hepatitis.BILLRUBIN.die, "pnorm")
ks.test(hepatitis.SGOT.die, "pnorm")
``` 

As we can notice none of the datasets follow a normal distribution.
Because we are running a test for 2 groups of a continuous variable 
with skewed data we choose the* Mann-Whitney U test to compare* the following 
groups: 

* hepatitis.BILLRUBIN.survived against hepatitis.BILLRUBIN.die
* hepatitis.SGOT.survived against hepatitis.SGOT.die

```{r, echo = TRUE}
wilcox.test(hepatitis.BILLRUBIN.die, hepatitis.BILLRUBIN.survived)
wilcox.test(hepatitis.SGOT.die, hepatitis.SGOT.survived)
``` 

In the first  hepatitis.BILLRUBIN.die and hepatitis.BILLRUBIN.survived
we have p-value of 6.521e-06 so we reject the null hypothesis concluding
that there is difference between then.

For the second case hepatitis.SGOT.die and hepatitis.SGOT.survived
we have a p-value = 0.2448 that is bigger than the $\alpha$ level of 0.05.
So we fail to reject the null hypothesis and say that there is no statistical
significant evidence that  hepatitis.SGOT.die and hepatitis.SGOT.survived are
different.

# Question 6 [2 marks]
Formulate a hypothesis by considering ALK PHOSPHATE levels and 
hepatitis histology by considering hepatitis dataset. Mention whether 
you accept or reject the hypothesis.

First we need to investigate

* H0 - 


## Answer 6

---------------------------------------------------


7. Investigate the following questions by considering the Hepatitis dataset.
[2 marks]
◦ Does Bilirubin level impact the Liver Firm?
◦ Are there any differences in steroid level and hepatitis histology?



